Tissue Enrichment Analysis for C. elegans Genomics

### 2016-06-29 tests/hgf_benchmarking/doc/tea_main_text.tex
minor changes to manuscript
added a p-value column
bc reviewer asked for it.

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. To interpret these long lists of genes, biologists need to abstract genes into concepts that are biologically relevant to form hypotheses about what is happening in the system. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed descriptions about the molecular, cellular or biochemical functions of any gene. For a given gene list, existing software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}. One area of biological significance that GO does not include is anatomy. One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}. Cells and tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from the literature. Here we provide a new framework that analyzes a user-input list for enrichment of specific cells and tissues. %As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.

Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene. %There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (pers. comm., H. Yu and P. Thomas). To reduce verbosity, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information. To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. Other pruning methods do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

%Our tool is available within WormBase and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.

\section*{Results}
\subsection*{Generating a Gene-Tissue Dictionary}
\subsubsection*{Reducing term redundancy through a similarity metric}
As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms. %We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic. For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing. 
For example, nematodes have a number of neurons that are placed symmetrically along the left/right body axis, and are functionally similar. These left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them%. We reasoned that when two tissues have almost identical annotations
, and therefore we cannot have statistical confidence in differentiating between them. As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters. %We refer to such sisters as `redundant'. 
To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}) that can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score. We referred to these two scoring criteria as `avg' and `any' respectively.

@@

\subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
By itself, each of these filters can reduce the number of nodes employed for analysis, but applying the filters in different orders removes different numbers of nodes (not all the filters are commutative). We chose to always execute annotation and similarity thresholding first, followed by the ceiling filter.
%If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets. Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any. On the other hand, applying the ceiling operator after trimming and redundancy filtering will result in greater numbers of nodes. We always applied the ceiling at the end. 
For validation (see below) we made a number of different dictionaries. The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations. After filtering, dictionary sizes ranged from 21 to a maximum of 460 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through the application of a few simple filters (see Table~\ref{tab:DictionarySpecs}). These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details).

@@

Once the p-values for each term have been calculated, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm~\cite{Benjamini1995}. FDR corrected p-values are called q-values. Genes that have a q-value less than a given alpha are considered significant. Our default setting is an alpha of 0.1.

Users input a gene list using any valid gene name for \emph{C.~elegans}. These names are processed into standard WormBase gene IDs (WBGene IDs). The program returns a table containing all the enriched terms and associated  information such as number of terms in gene list and expected number of terms.%with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
 Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values. The bars in the graph are sorted in ascending order of q-value and then in descending order of fold-change. Bars are colored for ease of viewing, and color does not convey information. Our software is implemented in an easy to use GUI (see Figure~\ref{fig:GUIresults}). Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID). In summary, each time the ontology annotations are updated, a new trimmed ontology is generated using our filters; in parallel, users can submit their gene lists through WormBase for testing, with results output in a number of formats (see Figure~\ref{fig:workflow}).

\subsection*{Validation of the algorithm and optimizing parameter selection}
We wanted to select a dictionary that included enough terms to be specific beyond the most basic %largest -> basic
 \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. Larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
%The phrase below doesn't seem useful
%To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}. These data sets are annotated gene lists derived from the corresponding Expression Cluster data in WormBase. Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set. To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.

@@

%dictionaries that work well
To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
%Work well
% TODO add WBbt codes!
Most results were comparable and expected. For some sets, all dictionaries performed well. For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays 
regardless of the dictionary used (see Table~\ref{tab:interagree}). On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `P$_4$',`P$_3$' and `P$_2$' were identified in addition to `Z2' and `Z3'. 
%Differences between dictionaries
We also queried an intestine precursor set~\cite{Spencer2011}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched. Both of these results capture specific aspects of \emph{C.~elegans} that are well known to developmental biologists.

@@

%Dictionary intra-agreement
We assessed the internal agreement of our tool by using independent gene sets that we expected to be enriched in the same tissues. We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD sets~\cite{Spencer2011, Smith2010}; and two  GABAergic sets~\cite{Spencer2011, Cinar2005}.%; two  pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two  intestinal gene sets~\cite{Spencer2011, Pauli2006}. 
Overall, the tool has good internal agreement. On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}), and most high-scoring terms were preserved between sets. %The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set showed no enriched terms in all dictionaries, so we did not consider them as informative for assessing internal agreement. 
All comparisons can be found online in our Github repository (see Availability of data and materials). Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion performed best, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary. As of this publication, the testable dictionary contains 261 terms.

\subsection*{Applying the tool}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology. Engelmann et al. exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct. Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes. As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}. These results also showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}). A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens. %One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes. \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile. 
Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.

@@

%Trimming discussion
We have also presented the first, to our knowledge, ontology trimming algorithm for biomedical ontologies. This algorithm, which is very easy to execute, places strong limits on the number of terms selected for testing. Due to the nature of all ontologies as hierarchical, acyclical graphs with term inheritance, term annotations are correlated along any given branch. This correlation reduces the benefits of including all terms for statistical analysis: for any given term along a branch, if that term passes significance, there is a high probability that many other terms along that branch will also pass significance. If the branch is enriched by random chance, error propagation along a branch means that many more false positives will follow. Thus, a researcher might be misled by the number of terms of correlated function and assign importance to this finding; the fact that the branching structure of GO amplifies false positive signals is a powerful argument for either reducing branch length or branch intracorrelation, or both. On the other hand, if a term is actually enriched, we argue that there is little benefit to presenting the user with additional terms along that branch. Instead, a user will benefit most from testing sparsely along the tree at a suitable specificity for hypothesis formation. Related terms of the same level should only be tested when there is sufficient annotation to differentiate, with statistical confidence, whether one term is enriched above the other. Our algorithm reduces branch length by identifying and removing nodes that are insufficiently annotated and parents that are likely to include sparse information.

We endeavoured to benchmark our tool well, but our analysis cannot address problems related to %identify good negative controls
 spurious term enrichment.
% Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results. For example, the embryonic germline precursor gene set had the term `AB', which refers to the anterior cell in the 2-cell stage \emph{C.~elegans} embryo, as the most enriched term in the dictionaries with cut off of 25 and 33. Is this an error, or does this hint at new biology? 
 Although we were unable to determine false-positive and false-negative rates, we do not believe this should deter scientists from using our tool. Rather, we encourage researchers to use our tool as a guide, integrating evidence from multiple sources to inform the most likely hypotheses. As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set. For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}. As time passes, we are certain the accuracy and power of this tool will improve thanks to the continuing efforts of the  worm research community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.

@@

@@

\section*{Availability of data and materials}
Our web implementation is available at \url{https://www.wormbase.org/enrichment/tea}. Our software can also be downloaded using Python's pip installer via the command

@@

% Figures describing algorithm
\begin{figure}[htbp]
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
   	 	\caption{}
    \label{fig:simdiagram}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
	      \caption{}
    \label{fig:trim_ends}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
    	\caption{}
    \label{fig:trim_roots}
  \end{subfigure}

@@

%results Figures - engelmann et al reanalysis
%labels are missing.
\begin{figure}[htbp]
    % \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{
	\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.~coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools. 
	\textbf{a)} pantherDB screenshot of results, sorted by p-value. Only top hits shown.
	\textbf{b)} TEA results, sorted by q-value (lowest on top) and fold-change. 
	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.
	}
	\label{fig:Dcon}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body

%%%%%%%%%%%%%%%%
%% Background %%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. To interpret these long lists of genes, biologists need to abstract genes into concepts that are biologically relevant to form hypotheses about what is happening in the system. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed descriptions about the molecular, cellular or biochemical functions of any gene. For a given gene list, existing software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009, Pathan2015}. One area of biological significance that GO does not include is anatomy. One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}. Such an ontology has been described previously for this organism~\cite{Lee2003}. Cells and tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from the literature. Here we provide a new framework that analyzes a user-input list for enrichment of specific cells and tissues.

Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene. DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (pers. comm., H. Yu and P. Thomas). To reduce verbosity, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information. To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. Other pruning methods do exist (see for example~\cite{Kim2007, Mambretti2013}), but the pruning is query-dependent or generates a brand new `brief ontology'. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

We have developed a tool that tests a user-provided list of genes for term enrichment using a nematode-specific tissue ontology. This ontology, which is not a module of Gene Ontology, is verbose. We trim our ontology using an algorithmic approach, outlined below, that reduces multiple hypothesis testing issues by limiting testing to terms that are well-annotated. The results are provided to the user in a GUI that includes a table of results and an automatically generated bar-chart. This software addresses a previously unmet need in the \emph{C.~elegans} community for a tool that reliably and specifically links gene expression with changes in specific cells, organs or tissues in the worm. 
\section*{Results}
\subsection*{Generating a Gene-Tissue Dictionary}
\subsubsection*{Reducing term redundancy through a similarity metric}
For our tool, we employ a previously generated cell and tissue ontology for \emph{C.~elegans}\cite{Lee2003}, which is maintained and curated by WormBase. This ontology contains thousands of anatomiy terms, but not every term is equally well-annotated. As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms. For example, nematodes have a number of neurons that are placed symmetrically along the left/right body axis, and are functionally similar. These left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them, and therefore we cannot have statistical confidence in differentiating between them. As a result, testing these sister terms provides no additional information compared with testing only the parent node to these sisters. To identify redundancy, we defined two possible similarity metrics (see \emph{Methods} section and Figure~\ref{fig:simdiagram}) that can be used to identify ontology sisters that have very high similarity between them. Intuitively, a set of sisters can be considered very similar if they share most gene annotations. Within a given set of sisters, we can calculate a similarity score for a single node by counting the number of unique annotations it contains and dividing by the total number of unique annotations in the sister set. Having assigned to each sister a similarity score, we can identify the \textbf{average} similarity score for this set of sisters, and if this average value exceeds a threshold, all of the sisters are removed from the ontology. An alternative method is check whether \textbf{any} of the scores exceeds a predetermined threshold, and if so remove this sister set from the ontology. We referred to these two scoring criteria as `\textbf{avg}' and `\textbf{any}' respectively.

@@

\subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
By itself, each of these filters can reduce the number of nodes employed for analysis, but applying the filters in different orders removes different numbers of nodes (not all the filters are commutative). We chose to always execute annotation and similarity thresholding first, followed by the ceiling filter.
%If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets. Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any. On the other hand, applying the ceiling operator after trimming and redundancy filtering will result in greater numbers of nodes. We always applied the ceiling at the end. 
For validation (see below) we made a number of different dictionaries. The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations. After filtering, dictionary sizes ranged from 21 to a maximum of 460 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through the application of a few simple filters (see Table~\ref{tab:DictionarySpecs}). These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details). Our trimming pipeline is applied with every new WormBase release. This ensures that the ontology database we are using remains up-to-date with regards to both addition or removal of specific terms as well as with regard to gene expression annotations.

@@

Once the p-values for each term have been calculated, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm~\cite{Benjamini1995}. FDR corrected p-values are called q-values. Genes that have a q-value less than a given alpha are considered significant. Our default setting is an alpha of 0.1, which is a standard threshold broadly agreed upon by the RNA-seq community (see for example \cite{Love2014}).

Users input a gene list using any valid gene name for \emph{C.~elegans}. These names are processed into standard WormBase gene IDs (WBGene IDs). The program returns a table containing all the enriched terms and associated  information such as number of terms in gene list and expected number of terms. Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values. The bars in the graph are sorted in ascending order of q-value and then in descending order of fold-change. Bars are colored for ease of viewing, and color does not convey information. Our software is implemented in an easy to use GUI (see Figure~\ref{fig:GUIresults}). Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID). In summary, each time the ontology annotations are updated, a new trimmed ontology is generated using our filters; in parallel, users can submit their gene lists through WormBase for testing, with results output in a number of formats (see Figure~\ref{fig:workflow}).

\subsection*{Validation of the algorithm and optimizing parameter selection}
We wanted to select a dictionary that included enough terms to be specific beyond the most basic \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. Larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power. To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}. These data sets are annotated gene lists derived from the corresponding Expression Cluster data in WormBase. Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set. To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.

@@

%dictionaries that work well
To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results. Most results were comparable and expected. For some sets, all dictionaries performed well. For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related regardless of the dictionary used (see Table~\ref{tab:interagree}). On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `P$_4$',`P$_3$' and `P$_2$' were identified in addition to `Z2' and `Z3'. 
%Differences between dictionaries
We also queried an intestine precursor set~\cite{Spencer2011}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched. Both of these results capture specific aspects of \emph{C.~elegans} that are well known to developmental biologists.

@@

%Dictionary intra-agreement
We assessed the internal agreement of our tool by using independent gene sets that we expected to be enriched in the same tissues. We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD sets~\cite{Spencer2011, Smith2010}; and two  GABAergic sets~\cite{Spencer2011, Cinar2005}. Overall, the tool has good internal agreement. On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}), and most high-scoring terms were preserved between sets. All comparisons can be found online in our Github repository (see Availability of data and materials). Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion performed best, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary. As of this publication, the testable dictionary contains 261 terms.

\subsection*{Applying the tool}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology. Engelmann et al. exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct. Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes. As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}. These results also showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}). A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens. Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.

@@

%Trimming discussion
We have also presented the first, to our knowledge, ontology trimming algorithm applied to biomedical ontologies. This algorithm, which is very easy to execute, places strong limits on the number of terms selected for testing. Due to the nature of all ontologies as hierarchical, acyclical graphs with term inheritance, term annotations are correlated along any given branch. This correlation reduces the benefits of including all terms for statistical analysis: for any given term along a branch, if that term passes significance, there is a high probability that many other terms along that branch will also pass significance. If the branch is enriched by random chance, error propagation along a branch means that many more false positives will follow. Thus, a researcher might be misled by the number of terms of correlated function and assign importance to this finding; the fact that the branching structure of GO amplifies false positive signals is a powerful argument for either reducing branch length or branch intracorrelation, or both. On the other hand, if a term is actually enriched, we argue that there is little benefit to presenting the user with additional terms along that branch. Instead, a user will benefit most from testing sparsely along the tree at a suitable specificity for hypothesis formation. Related terms of the same level should only be tested when there is sufficient annotation to differentiate, with statistical confidence, whether one term is enriched above the other. Our algorithm reduces branch length by identifying and removing nodes that are insufficiently annotated and parents that are likely to include sparse information.

We endeavoured to benchmark our tool well, but our analysis cannot address problems related to spurious term enrichment. Although we were unable to determine false-positive and false-negative rates, we do not believe this should deter scientists from using our tool. Rather, we encourage researchers to use our tool as a guide, integrating evidence from multiple sources to inform the most likely hypotheses. As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set. For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}. As time passes, we are certain the accuracy and power of this tool will improve thanks to the continuing efforts of the  worm research community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.

@@

Since nodes can have multiple parents (and therefore multiple sister sets), a complete set of similarity scores was calculated before trimming the ontology, and nodes were removed from the ontology if they exceeded the similarity threshold at least once in any comparison.

@@

\section*{Availability of data and materials}
Our web implementation is available at \url{http://www.wormbase.org/tools/enrichment/tea/tea.cgi}. Our software can also be downloaded using Python's pip installer via the command

@@

% Figures describing algorithm
\begin{figure}[htbp]
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/Redundancy_Trimming_Ontology.pdf}
   	 	\caption{}
    \label{fig:simdiagram}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/Floor_Trimming_Ontogeny.pdf}
	      \caption{}
    \label{fig:trim_ends}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    % \includegraphics[width=\textwidth]{figures/Ceiling_Trimming_Ontology.pdf}
    	\caption{}
    \label{fig:trim_roots}
  \end{subfigure}

@@

%results Figures - engelmann et al reanalysis
%labels are missing.
\begin{figure}[htbp]
    % \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.pdf}
	\captionsetup{width= 0.95\textwidth}
  	\caption{
	\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.~coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools. 
	\textbf{a)} pantherDB screenshot of results, sorted by p-value. Only top hits shown.
	\textbf{b)} TEA results, sorted by q-value (lowest on top) and fold-change. 
	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.
	}
	\label{fig:Dcon}
\end{figure}
